---
title: "Analysis of serve-flation"
author: "Andrew Tredennick"
format: html
---

# Libraries and settings

```{r libraries}
library(tidyverse)
library(here)
library(arrow)
library(brms)
library(tidybayes)
library(ggridges)
library(showtext)
```

```{r settings}
showtext_auto()  # automatically use showtext for rendering
font_add_google("Oswald", "myfont")

theme_custom <- theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "myfont", color = "#333333"),
    plot.background = element_rect(fill = "#F7F7F7", color = NA),
    panel.background = element_rect(fill = "#F7F7F7", color = NA),
    legend.position = "top",  
    legend.direction = "horizontal",  
    legend.justification = "left", 
    legend.box = "horizontal",
    legend.box.just = "left",  
    legend.background = element_blank(),
    legend.box.background = element_blank()
  )
theme_set(theme_custom)

# from On Cloud (ON) color pallette: https://www.on.com
color_map2 <- c(Iga = "#FC63CE", Opponents = "#27414F")
color_map4 <- c("#D1D1D1", "#B8A89D", "#8E7F6C", "#707070") 
```

# Data

## Function to pre-process serve data
This function takes the full serve dataset and subsets to a player of interest based on their surname (last name).
The player of interest is specifically identified as the server when the record is for their serve.
All other servers (every one of their opponents in the dataset) are collectively defined as "Opponents."

```{r data-fxn}
process_player_data <- function(data, surname, idname) {
  # filter data
  data <- data |>
    filter(
      (str_detect(player1, surname) | str_detect(player2, surname)) &
      Speed_KMH > 0 & !is.na(Speed_KMH)
    ) |>
    mutate(
      PointServer = case_when(
        PointServer == 1 ~ player1,
        PointServer == 2 ~ player2,
        TRUE ~ NA_character_
      )
    )
  
  # transform and select relevant columns
  data <- data |>
    mutate(
      server = case_when(
        str_detect(PointServer, surname) ~ idname,
        TRUE ~ "Opponents"
      )
    ) |>
    filter(year > 2020, slam %in% c("usopen", "wimbledon")) |>
    dplyr::select(
      match_id, slam, year,
      server, speed_kmh = Speed_KMH
    )
  
  # format columns for analysis
  data <- data |>
    mutate(
      year = factor(year),
      server = factor(server, levels = c(idname, "Opponents"))
    )
  
  return(data)
}

```

## Load data and process
```{r data}
serve_data <- read_parquet(here("data", "serve_data.parquet"))
iga_data <- process_player_data(serve_data, "Swiatek", "Iga")
```

## Plot serve speeds over time by slam
```{r wta}
facet_map <- c(usopen = "US Open", wimbledon = "Wimbledon")
eda <- ggplot(iga_data,
       aes(
         x = factor(year),
         y = speed_kmh,
         fill = server,
         color = server
       )) +
  geom_dotplot(
    binaxis = 'y',
    binwidth = 2,
    dotsize = 0.5,
    stackdir = 'center',
    position = position_dodge(0.5),
    alpha = 0.3
  ) +
  geom_boxplot(position = position_dodge(0.5),
               alpha = 0.2,
               width = 0.3,
               outlier.shape = NA) +
  facet_wrap( ~ slam, ncol = 1, labeller = as_labeller(facet_map)) +
  scale_color_manual(values = color_map2, name = NULL) +
  scale_fill_manual(values = color_map2, name = NULL) +
  labs(x = "Year", y = "Serve speed (km/h)")

ggsave(here("figures", "speed-data.pdf"), plot = eda, height = 5, width = 6)
```


# Model

## Fit regression model
Fit a simple model of serve speeds with a year-by-server interaction and an additive effect os slam (in case there is systematic bias due to technology at each venue).
Using a weuibull to maybe capture some skew in observations.
```{r fit}
fit <- brm(speed_kmh ~ server*year + slam, 
           data = iga_data, 
           family = weibull(), 
           chains = 4, 
           cores = 4, 
           warmup = 1000, 
           iter = 2000, 
           seed = 20241121)
```

## Generate model expectations for inference
```{r preds}
newdata <- tibble(server = unique(iga_data$server)) |>
  crossing(tibble(year = unique(iga_data$year))) |>
  crossing(tibble(slam = unique(iga_data$slam)))

preds <- add_epred_draws(newdata, object = fit)
```

# Model inference

## Is Iga's serve getting faster?
```{r}
iga_over_time <- preds |>
  ungroup () |>
  dplyr::select(server, year, .epred, .draw) |>
  group_by(server, year, .draw) |>
  summarise(speed = mean(.epred), .groups = "drop") |>
  ungroup() |>
  filter(server == "Iga") |>
  pivot_wider(names_from = year, values_from = speed) |>
  pivot_longer(cols = `2022`:`2024`) |>
  mutate(delta = value - `2021`)

iga_time_plot <- ggplot(data = iga_over_time, aes(x = delta, y = name, fill = name)) +
  geom_vline(aes(xintercept = 0), linetype = 2) +
  geom_density_ridges(alpha = 0.75,
                      color = NA,
                      bandwidth = 0.5) +
  scale_fill_manual(values = color_map4[2:4]) +
  coord_cartesian(xlim = c(-5, 10)) +
  labs(x = "Iga's serve speed increase/decrease relative to 2021 (km/h)", 
       y = NULL, 
       caption = "\n\nPosterior distributions averaged over\nUS Open and Wimbledon serve speeds.") +
  guides(fill = "none") 

ggsave(here("figures", "iga_over_time.pdf"), iga_time_plot, width = 6, height = 4)
```

Yes, except for dip in 2022.

## Are Iga's opponent's serves getting faster?
```{r}
opp_over_time <- preds |>
  ungroup () |>
  dplyr::select(server, year, .epred, .draw) |>
  group_by(server, year, .draw) |>
  summarise(speed = mean(.epred), .groups = "drop") |>
  ungroup() |>
  filter(server == "Opponents") |>
  pivot_wider(names_from = year, values_from = speed) |>
  pivot_longer(cols = `2022`:`2024`) |>
  mutate(delta = value - `2021`)

opp_time_plot <- ggplot(data = opp_over_time, aes(x = delta, y = name, fill = name)) +
  geom_vline(aes(xintercept = 0), linetype = 2) +
  geom_density_ridges(alpha = 0.75, color = NA, bandwidth = 0.5) +
  scale_fill_manual(values = color_map4[2:4]) +
  coord_cartesian(xlim = c(-5, 10)) +
  labs(
    x = "Opponents's serve speed increase/decrease relative to 2021 (km/h)",
    y = NULL,
    caption = "\n\nPosterior distributions averaged over\nUS Open and Wimbledon serve speeds."
  ) +
  guides(fill = "none")

ggsave(here("figures", "opp_over_time.pdf"), opp_time_plot, width = 6, height = 4)
```

Yes, but really just a step-change from 2021, then pretty constant.

## Is Iga keeping up with serve-flation?
```{r}
deltas <- preds |>
  ungroup () |>
  dplyr::select(server, year, .epred, .draw) |>
  group_by(server, year, .draw) |>
  summarise(speed = mean(.epred), .groups = "drop") |>
  ungroup() |>
  pivot_wider(names_from = server, values_from = speed) |>
  mutate(delta = Iga - Opponents)

delta_plot <- ggplot(deltas, aes(x = delta, y = year, fill = year, group = year)) +
  geom_vline(aes(xintercept = 0), linetype = 2) +
  geom_density_ridges(alpha = 0.75, color = NA, bandwidth = 0.5) +
  scale_fill_manual(values = color_map4) +
  labs(
    x = "Iga's speed advantage (km/h)",
    y = NULL,
    caption = "\n\nPosterior distributions averaged over\nUS Open and Wimbledon serve speeds."
  ) +
  guides(fill = "none") 
ggsave(here("figures", "deltas.pdf"), plot = delta_plot, width = 6, height = 4)
```

Yes, Iga is actually gaining ground each year.
